---
title: "Pomacentrid differential expression analysis"
author: "Eirlys Tysall"
date: "2023-02-24"
output: html_document
---

```{r load-packages, include=FALSE}
#load packages
suppressPackageStartupMessages({
  library(data.table)
  library(tidyverse)
  library(tximport)
  library(markdown)
  library(DESeq2)
  library(SummarizedExperiment)
  library(resrepo)
})
```

```{r load-functions}
#load functions
#make tx2gene object function
source(path_resrepo("/code/r_functions/make_tx2gene.R"))

#join multiple tximport outputs function
source(path_resrepo("/code/r_functions/split_combine_rejoin.R"))
```

```{r read-in-data, include=FALSE}
#read in atrilobata annotations
atri_annotations <- read_tsv(path_resrepo("/data/intermediate/p08_differential_expression/s00_final_annotation/pom/atrilobata_final_annotations.tsv"))

# read in multilineata annotations
multi_annotations <- read_tsv(
  path_resrepo("/data/intermediate/p08_differential_expression/s00_final_annotation/pom/multilineata_final_annotations.tsv"),
  col_types = cols(
    evalue = col_character(),  # specify columns 3 and 4 as character
    score = col_character()
    # Add other column types as needed
  )
)

# read in cyanea annotations
cyanea_annotations <- read_tsv(path_resrepo("/data/intermediate/p08_differential_expression/s00_final_annotation/pom/cyanea_final_annotations.tsv"))

#salmon quant results directory
salmon_dir <- path_resrepo(("/data/intermediate/s07_salmon/pom/final"))

#read in meta data info
meta <- read_csv(file.path(salmon_dir, "meta_pom.txt"), col_types=cols(.default=col_factor()))

#salmon quant.sf
#atri
a_quant <- read_tsv(file.path(salmon_dir,"S24/quant.sf")) 
b_quant <- read_tsv(file.path(salmon_dir,"S25/quant.sf")) 
c_quant <- read_tsv(file.path(salmon_dir,"S26/quant.sf")) 
d_quant <- read_tsv(file.path(salmon_dir,"S27/quant.sf")) 
#multi
e_quant <- read_tsv(file.path(salmon_dir,"S1/quant.sf"))
f_quant <- read_tsv(file.path(salmon_dir,"S2/quant.sf"))
g_quant <- read_tsv(file.path(salmon_dir,"S3/quant.sf"))
h_quant <- read_tsv(file.path(salmon_dir,"S4/quant.sf"))
#cyanea
i_quant <- read_tsv(file.path(salmon_dir,"S20/quant.sf"))
j_quant <- read_tsv(file.path(salmon_dir,"S21/quant.sf"))
k_quant <- read_tsv(file.path(salmon_dir,"S22/quant.sf"))
l_quant <- read_tsv(file.path(salmon_dir,"S23/quant.sf"))

```

```{r diff-exp}
##RUNS FULL DIFFERENTIAL EXPRESSION ANALYSIS

#extract atri
meta_1 <- meta %>%
   dplyr::slice(1:4)

#location of all salmon quant.sf files
salmonfiles_1 <- file.path(salmon_dir, meta_1$Name,"quant.sf")
names(salmonfiles_1) <- meta_1$Name

#check all files named in meta_info.txt present
stopifnot(all(file.exists(salmonfiles_1)))

###

#extract multi
meta_2 <- meta %>%
   dplyr::slice(5:8)

#location of all salmon quant.sf files
salmonfiles_2 <- file.path(salmon_dir, meta_2$Name,"quant.sf")
names(salmonfiles_2) <- meta_2$Name

#check all files named in meta_info.txt present
stopifnot(all(file.exists(salmonfiles_2)))

###

#extract cyanea
meta_3 <- meta %>%
   dplyr::slice(-(1:8))

meta_1

#location of all salmon quant.sf files
salmonfiles_3 <- file.path(salmon_dir, meta_3$Name,"quant.sf")
names(salmonfiles_3) <- meta_3$Name

#check all files named in meta_info.txt present
stopifnot(all(file.exists(salmonfiles_3)))

##############################################

#atrii
#create tx2gene object

atri_tx2gene_object <- make_tx2gene(eggnog_annotation_file = atri_annotations,
                                   salmon_quant_file_1 = a_quant,
                                   salmon_quant_file_2 = b_quant,
                                   salmon_quant_file_3 = c_quant, 
                                   salmon_quant_file_4 = d_quant)

#atri
#use tx2gene object to collapse trinity isoforms to gene level annotations
sg_atri <- tximport(files = salmonfiles_1, 
                   type = "salmon", 
                   tx2gene = atri_tx2gene_object,
                   ignoreAfterBar = TRUE)

####################################################

#multi
#create tx2gene object 
multi_tx2gene_object <- make_tx2gene(eggnog_annotation_file = multi_annotations,
                                   salmon_quant_file_1 = e_quant,
                                   salmon_quant_file_2 = f_quant,
                                   salmon_quant_file_3 = g_quant, 
                                   salmon_quant_file_4 = h_quant)


#multi
#use tx2gene object to collapse trinity isoforms to gene level annotations
sg_multi <- tximport(files = salmonfiles_2, 
                   type = "salmon", 
                   tx2gene = multi_tx2gene_object, 
                   ignoreAfterBar = TRUE)


##################################################

#cyanea
#create tx2gene object 
cyanea_tx2gene_object <- make_tx2gene(eggnog_annotation_file = cyanea_annotations,
                                   salmon_quant_file_1 = i_quant,
                                   salmon_quant_file_2 = j_quant,
                                   salmon_quant_file_3 = k_quant, 
                                   salmon_quant_file_4 = l_quant)




#cyanea
#use tx2gene object to collapse trinity isoforms to gene level annotations
sg_cyanea <- tximport(files = salmonfiles_3, 
                   type = "salmon", 
                   tx2gene = cyanea_tx2gene_object, 
                   ignoreAfterBar = TRUE)

#####################################################

#tximport output has 4 elements: counts, abundance, length, countsFromAbundance
#Here I extract each equivalent element from colonus and furcifer tximport output 
#and join them using matching row names 

atri_counts <- rownames_to_column(as.data.frame(sg_atri$counts))
atri_abd <- rownames_to_column(as.data.frame(sg_atri$abundance))
atri_len <- rownames_to_column(as.data.frame(sg_atri$length))

multi_counts <- rownames_to_column(as.data.frame(sg_multi$counts))
multi_abd <- rownames_to_column(as.data.frame(sg_multi$abundance))
multi_len <- rownames_to_column(as.data.frame(sg_multi$length))

cyanea_counts <- rownames_to_column(as.data.frame(sg_cyanea$counts))
cyanea_abd <- rownames_to_column(as.data.frame(sg_cyanea$abundance))
cyanea_len <- rownames_to_column(as.data.frame(sg_cyanea$length))

##add in 0 for counts 
atri_multi_counts <- full_join(atri_counts, multi_counts)

all_counts<- full_join(atri_multi_counts, cyanea_counts)

all_counts_na <- all_counts[!complete.cases(all_counts), ] #2959

all_counts[is.na(all_counts)] <- 0

##add in 0 for abundance

atri_multi_abd <- full_join(atri_abd, multi_abd)

all_abd<- full_join(atri_multi_abd, cyanea_abd)

all_abd_na <- all_abd[!complete.cases(all_abd), ] #2959

all_abd[is.na(all_abd)] <- 0

##add in average for length

atri_multi_len <- full_join(atri_len, multi_len)

all_len<- full_join(atri_multi_len, cyanea_len)

all_len_na <- all_len[!complete.cases(all_len), ] #2959


#function for length average
replace_na_with_mean <- function(row) {
  mean_value <- mean(row, na.rm = TRUE)  # Calculate mean excluding NAs
  row[is.na(row)] <- mean_value           # Replace NAs with mean
  return(row)
}

# Apply the function to each row
all_len[, -1] <- t(apply(all_len[, -1], 1, replace_na_with_mean))

all_counts_m <- all_counts %>% column_to_rownames() %>% as.matrix()
all_abd_m <- all_abd %>% column_to_rownames() %>% as.matrix()
all_len_m <- all_len %>% column_to_rownames() %>% as.matrix()

all_sg <- list()
  
  #Add each matriix to list 
  all_sg[["abundance"]] <- all_abd_m
  all_sg[["counts"]] <- all_counts_m
  all_sg[["length"]] <- all_len_m
  all_sg[["countsFromAbundance"]] <- "no"
  
  
sg <- all_sg


###############################################


#creates _DESeqDataSet_
ds_se <- DESeq2::DESeqDataSetFromTximport(txi = sg, 
                                  colData = meta,
                                  design = ~ Species)

#################################################

#runs DESeq diff expression analysis
ds_se <- DESeq2::DESeq(ds_se)
```

```{r subset-results}
res_gem_filt <- DESeq2::results(ds_se, 
                               name="Species_A.multilineata_vs_A.atrilobata", 
                               lfcThreshold = 0.5, 
                               alpha = 0.05)

gene_code <- c("49XM4","49QHR","49VMA","4A6WB","49XFU1","49XFU2","49WBX","49TYN",
            "4A0DI","49UVW","4A127","49TK5","49WJ7","49UFF","49S7R","4A2MR",
            "49YH8","49VB3","49SJD","49T5U","49R5D","49WZN","49S06","49PYM","PARIE",
            "4A27E","49XC4","4A5MG","4A13M","49V8D","4A0TV","49XUM","49VYF1",
            "49VYF2","49VUD","4A0N0","49XUW1","49XUW2","49QHG","49WTJ","49YV11",
            "49YV12","49SUB","49R8Z","49XPK","49Y02","49TP6","49XM9","4A6Y0",
            "49UXN","4A6E6","49Z3E",
            "49S3W","49SQ7","49R2B","4A582","4A63P","49VV8","49ZE6","49UB5",
            "4A1XJ","49VYA","49UJA","49UMK","49WSA","49XSJ","49RM7","4A4PS",
            "4A4VB","49VSK","49YFA","49ZMT","4A1II","49QNB","49VVR","49UCK",
            "49RFK","4A44T","4A4FK","49WVI","49YJN","4A0K4","49WUQ")

gene_name <- c("Exo-rho","opn3","opn4a","opn4b","opn4xa","opn4xb","opn5","opn6a","opn7a",
           "opn7b","opn7d","opn8a","opn8b","opn8c","opn9","parapinopsin-like","rgra",
           "rgrb","rrh","tmtops2b","tmtops3a","tmtopsa","tmtopsb","VAL-opsin-a","parietopsin",
           "aanat1","arntl","ciart","clock a","clock b","cry-dash","cry1-like","cry1a",
           "cry1b","cry2","cry5","nocta","noctb","npas2","nr1d1","nr1d2a",
           "nr1d2b","per1","per2","per3","rora","rorb","timeless", "csnk1da",
           "csnk1db","csnk1ea","csnk1eb",
           "RCVRN","GRK1","GRK7","SAG","ARR3","CNGA1","CNGA3","CNGB1",
           "CNGB3","gnat1","gnat2","GNB1","GNB3","GNB5a","GNB5b","GNGT1",
           "GNGT2","guca1a","guca1b","guca1c","gucy2d","gucy2f","pde6a","pde6b",
           "pde6c","pde6g","pde6h","RGS9","RGS9BP","slc24a1","slc24a2")

df <- data.frame(
  gene = gene_code,
  gene_name = gene_name
)

#subset results
res_gem_sub <- res_gem_filt[row.names(res_gem_filt) %in% gene_code, ]
res_gem_sub$padj <- p.adjust(res_gem_sub$pvalue, method="BH")

res_gem_df <- rownames_to_column(as.data.frame(res_gem_sub))
colnames(res_gem_df)[1]<- "gene"

full <-full_join(df, res_gem_df, by = "gene")
```

```{r}
#any(is.na(rownames(res_gem_sub)))

# If there are missing row names, you can assign new row names
#rownames(res_gem_sub) <- seq_len(nrow(res_gem_sub))

# Now, try to convert to a data frame with row names as a column
#res_gem_df <- rownames_to_column(as.data.frame(res_gem_sub))
```

```{r save-subset}
#save subset results 
writexl::write_xlsx(full, path = path_resrepo("data/intermediate/p08_differential_expression/s02_differential_expression/pom/pom_subset_filt.xlsx"))
```


##SAVE DATA OBJECTS##

```{r vst-transform, message=FALSE}
#Variance stabilising transformation (VST), a heuristic that transforms the 
# variance so it becomes independent of mean
vsd <- DESeq2::vst(ds_se)

# save vsd data for use in downstream analysis 
save(vsd, file = path_resrepo("/data/intermediate/p08_differential_expression/s02_differential_expression/pom/data_objects/pom_vsd"))
```

```{r save-sg}
#save sg data for use in opsin expression analysis 
save(sg, file = path_resrepo("/data/intermediate/p08_differential_expression/s02_differential_expression/pom/data_objects/pom_sg"))

#ind species sg
#atri
save(sg_atri, file = path_resrepo("/data/intermediate/p08_differential_expression/s02_differential_expression/pom/data_objects/atri_sg"))

#multi
save(sg_multi, file = path_resrepo("/data/intermediate/p08_differential_expression/s02_differential_expression/pom/data_objects/multi_sg"))

#atri
save(sg_cyanea, file = path_resrepo("/data/intermediate/p08_differential_expression/s02_differential_expression/pom/data_objects/cyanea_sg"))
```

```{r save-ds_se}
#save DESeqDataSet for downstream analysis
save(ds_se, file = path_resrepo("/data/intermediate/p08_differential_expression/s02_differential_expression/pom/data_objects/pom_ds_se"))
```

##RUN AND SAVE RESULTS OBJECTS -- FULL RESULTS OF ALL GENES ##

### ATRILOBATA vs MULTI

```{r gem-results}
#make res object with comparison want
res_gem <- DESeq2::results(ds_se, name = "Species_A.multilineata_vs_A.atrilobata")
```

```{r gem_save}
#save geminate comparison
res_gem <- as.data.frame(res_gem)
res_gem <- setDT(res_gem, keep.rownames = "gene")

write.csv(res_gem, file = path_resrepo("/data/intermediate/p08_differential_expression/s02_differential_expression/pom/unfiltered_results/pom_res_gem.csv"), row.names = FALSE)
```

### CYANEA vs ATRILOBATA

```{r cya-atri-results}
#make res object with comparison want
res_cya_atri <- DESeq2::results(ds_se, name = "Species_A.cyanea_vs_A.atrilobata")
```

```{r cya-atri-save}
#save results of cyanea atri comparison 
res_cya_atri <- as.data.frame(res_cya_atri)
res_cya_atri <- setDT(res_cya_atri, keep.rownames = "gene")

write.csv(res_cya_atri, file = path_resrepo("/data/intermediate/p08_differential_expression/s02_differential_expression/pom/unfiltered_results/pom_res_cya_atri.csv"),
          row.names = FALSE)

```

### CYANEA vs MULTI

```{r cya-mul-results}
#change levels before run in order to get cyanea vs multi comparison
ds_se1 <- ds_se

#levels
ds_se1$Species <- relevel(ds_se1$Species, ref = "A.multilineata")

#runs DESeq diff expression analysis
ds_se1 <- DESeq2::DESeq(ds_se1)

#check levels
DESeq2::resultsNames(ds_se1)

#make res object with comparison want
res_cya_multi <- DESeq2::results(ds_se1, name = "Species_A.cyanea_vs_A.multilineata")
```

```{r cya-mul-save}
#save results of cyanea multi comparison 
res_cya_multi <- as.data.frame(res_cya_multi)
res_cya_multi <- setDT(res_cya_multi, keep.rownames = "gene")

write.csv(res_cya_multi, file = path_resrepo("/data/intermediate/p08_differential_expression/s02_differential_expression/pom/unfiltered_results/pom_res_cya_multi.csv"), row.names = FALSE)
```

###SAVE WITH FILTERING AS PER SCHURCH PAPER

```{r filtering}
#filter results with minimum lfc = 0.5 and padj = 0.01
res_gem_filt <- DESeq2::results(ds_se, 
                               name="Species_A.multilineata_vs_A.atrilobata", 
                               lfcThreshold = 0.5, 
                               alpha = 0.05)
#how many sig overall
summary(res_gem_filt)
table(res_gem_filt$padj < 0.05)

res_cya_atri_filt <- DESeq2::results(ds_se, 
                               name="Species_A.cyanea_vs_A.atrilobata", 
                               lfcThreshold = 0.5, 
                               alpha = 0.05)

res_cya_multi_filt <- DESeq2::results(ds_se1, 
                               name="Species_A.cyanea_vs_A.multilineata", 
                               lfcThreshold = 0.5, 
                               alpha = 0.05)

```


```{r save-filtered}
#save filtered results
#geminates
res_gem_filt <- as.data.frame(res_gem_filt)
res_gem_filt <- setDT(res_gem_filt, keep.rownames = "gene")
write.csv(res_gem_filt, file = path_resrepo("data/intermediate/p08_differential_expression/s02_differential_expression/pom/filtered_results/pom_res_gem_filt.csv"), row.names = FALSE)

#cyanea vs atrilobata
res_cya_atri_filt <- as.data.frame(res_cya_atri_filt)
res_cya_atri_filt <- setDT(res_cya_atri_filt, keep.rownames = "gene")
write.csv(res_cya_atri_filt, file = path_resrepo("data/intermediate/p08_differential_expression/s02_differential_expression/pom/filtered_results/pom_res_cya_atri_filt.csv"), row.names = FALSE)

#cyanea vs multilineata
res_cya_multi_filt <- as.data.frame(res_cya_multi_filt)
res_cya_multi_filt <- setDT(res_cya_multi_filt, keep.rownames = "gene")
write.csv(res_cya_multi_filt, file = path_resrepo("data/intermediate/p08_differential_expression/s02_differential_expression/pom/filtered_results/pom_res_cya_multi_filt.csv"), row.names = FALSE)

```

```{r session-info}
#Session Info
sessionInfo()
```

R version 4.3.2 (2023-10-31)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 20.04.6 LTS

Matrix products: default
BLAS/LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/libopenblasp-r0.3.8.so;  LAPACK version 3.9.0

locale:
 [1] LC_CTYPE=en_GB.UTF-8       LC_NUMERIC=C               LC_TIME=en_GB.UTF-8        LC_COLLATE=en_GB.UTF-8    
 [5] LC_MONETARY=en_GB.UTF-8    LC_MESSAGES=en_GB.UTF-8    LC_PAPER=en_GB.UTF-8       LC_NAME=C                 
 [9] LC_ADDRESS=C               LC_TELEPHONE=C             LC_MEASUREMENT=en_GB.UTF-8 LC_IDENTIFICATION=C       

time zone: Europe/London
tzcode source: system (glibc)

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods   base     

loaded via a namespace (and not attached):
 [1] utf8_1.2.4                  generics_0.1.3              SparseArray_1.2.4           bitops_1.0-7               
 [5] DESeq2_1.42.1               lattice_0.21-9              digest_0.6.35               magrittr_2.0.3             
 [9] evaluate_0.23               grid_4.3.2                  fastmap_1.2.0               Matrix_1.6-1.1             
[13] writexl_1.5.0               GenomeInfoDb_1.38.8         fansi_1.0.6                 scales_1.3.0               
[17] codetools_0.2-19            abind_1.4-5                 cli_3.6.2                   rlang_1.1.3                
[21] crayon_1.5.2                XVector_0.42.0              Biobase_2.62.0              munsell_0.5.1              
[25] DelayedArray_0.28.0         yaml_2.3.8                  S4Arrays_1.2.1              tools_4.3.2                
[29] parallel_4.3.2              BiocParallel_1.36.0         dplyr_1.1.4                 colorspace_2.1-0           
[33] ggplot2_3.5.1               locfit_1.5-9.9              GenomeInfoDbData_1.2.11     SummarizedExperiment_1.32.0
[37] BiocGenerics_0.48.1         vctrs_0.6.5                 R6_2.5.1                    matrixStats_1.3.0          
[41] stats4_4.3.2                lifecycle_1.0.4             zlibbioc_1.48.2             S4Vectors_0.40.2           
[45] IRanges_2.36.0              pkgconfig_2.0.3             gtable_0.3.5                pillar_1.9.0               
[49] glue_1.7.0                  Rcpp_1.0.12                 xfun_0.44                   tibble_3.2.1               
[53] GenomicRanges_1.54.1        tidyselect_1.2.1            rstudioapi_0.16.0           MatrixGenerics_1.14.0      
[57] knitr_1.46                  htmltools_0.5.8.1           rmarkdown_2.26              compiler_4.3.2             
[61] RCurl_1.98-1.14            


